---
title: "Getting Started with tagbio.R"
author: "J Ireland, j@tag.bio"
date: "January 14, 2020"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{getting-started-with-tagbio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## 1 Overview of Using tag.bio with R

The `tagbio.R` package allows users to interact with a tag.bio flux capacitor
(FC) using native R commands.  Communication can happen in both directions: requesting
data from tag.bio to R and also sending results and visualizations back to a running
tag.bio instance.  

### 1.3 Calling tag.bio from R

In the simplest use case, a user may wish to retrieve data from a tag.bio FC
to use in their R application or notebook.  In this scenario, the FC is being used as a 
rich, queryable data source, returning tabular results much like they would
get through a database call or csv file upload.  The interaction is one way
with data being pulled from the FC, but no results being returned back to 
tag.bio.  The user is doing almost all of their work from the R environment.
The `tagbio.R` package facilitates this interaction, performing API calls and
formatting results as R data.frame (tibble) objects.

Even if the user's end goal is to run custom R methods from a tag.bio protocol,
method development will typically use this read-only interaction mode until the code
is deemed ready for use.  Once the R method is working in this mode, it is 
straight-forward to then have it be called directly from the tag.bio protocol as 
described below.

### 1.2 Calling R from tag.bio

Tag.bio can leverage custom user methods written in the R language as well as make use
of the extensive universe of R libraries and visualizations.  Tag.bio allows users to add these 
methods to their FC instances through an easy configuration step.  The developer's task
is to write a custom method which takes as input a `TagbioData` object and returns a `TagbioResult`
object.  The returned results can be PDFs, images or data frames which tag.bio can then incorporate
into the protocol workflow.

## 2 Installation

For now, the `tagbio.R` package is in beta and must be installed manually.  Make sure you have a recent
version of R installed (3.5 or higher).  Clone the tagbio.R repository, install the R dependencies and
finally install `tagbio.R` as shown below.

```

$ git clone https://jireland@bitbucket.org/cloudfolks/r-utils.git
$ Rscript -e 'install.packages(c("tidyverse", "argparse", "httr",
  "rjson", "qpdf"), repos="https://cran.rstudio.com")'
$ R CMD INSTALL r-utils/tagbio_0.1.0.tar.gz

```


## 3 Example: Calling tag.bio from R

We first create a FC instance which tells R the name and location of the tag.bio FC we wish to communicate with.  If the
FC was private, we would also need to provide a valid API key.  For now, we will work with `fc-hnsc`, an open FC hosted by tag.bio.
This FC contains multi-omics data from 530 head and neck cancer patients from the TCGA project.   

```{r create_fc, echo=TRUE, message=F, warning=F}

library(tagbio)
library(tidyverse)

# set up connection to the FC
fc_hnsc <- FC(name = "fc-hnsc", 
              url = "https://fc-genesig-gateway-azure.dev.tag.bio/fc-svc/fc-hnsc/")

```


After creating the FC, we must specify an existing protocol from the FC that we wish to pull data from.  Through the protocol instance
we can specify which data attributes we want to return as well as limit the returned results to a subset of the full protocol results.  

In this example, we will look at the expression of a set of leukocyte infiltration genes (Wolf, 2014).  We specify in the protocol instance
the gene symbol of the signature genes and also request tag.bio to return the TCGA sample identifiers. 


```{r create_protocol, echo=TRUE, message=F, warning=F}

# leukocyte infiltration signature genes
leuk_genes <- c("CD19", "CD37", "CD3D", "CD3E", "CD3G", "CD247", "CD79A", 
    "CD79B", "CD8A",  "CD14", "LCK", "LTB", "MS4A1")

# define the protocol instance
prot_inst <- ProtocolInstance(name = "download", 
    arguments = list(expression = leuk_genes,
                     clinical_categorical_variables = c("clinical.SAMPLE_ID")))

```


Up to this point, we have not yet communicated with tagbio.R.  We do this now through the `getTagData` function.  This function requires the
FC and protocol we just defined.  This function is the primary way we retrieve data from tagbio.R.  It returns a `TagbioData` object as we will 
see shortly.


```{r get_data, echo=TRUE, message=F, warning=F}

# get data from tagbio
tag_data <- getTagData(fc_hnsc, prot_inst) 
tag_data@data.frame[c(1:5),c(1:4)]

```


As you can see, the `TagbioData` object contains a data frame (technically, a tidyverse `tibble`) of our results.  `TagbioData` objects can also
return other data such as parameter "pass throughs".  These are typically used more when R is being invoked from tag.bio.  We'll discuss them 
more later.


The data frame that is returned by tag.bio will have no explicit row names.  Column names will contain the data type as well as attribute name, 
such as gene symbol.  This naming avoids potential name conflicts if, for example, two types of data were requested for the same gene.  The 
`tagbio` library provides a helper function `getDataFrame` which can be used to get data of a single type and can also add row names. 


```{r get_expr_array, echo=TRUE, message=F, warning=F}

# get a table of the gene expression results
exp_data <- getDataFrame(tag_data, data_type = "Expression", 
                         row_name ="clinical.SAMPLE_ID")
exp_data[c(1:5), c(1:5)]

```

The `exp_data` returned by `getDataFrame` is a matrix of normalized and log2 scaled gene expression values for our leukocyte signature genes.
Rows are labeled by the TCGA sample identifier.  The data is now ready for further downstream analysis and visualizations leveraging all of 
R's functionality and extensive libraries.  Here we will perform an unsupervised, hierarchical bi-clustering and visualize the results with a
heatmap. 


```{r draw_heat_map, echo=TRUE, message=F, warning=F}

library(gplots)

# remove NA's from expression matrix
exp_data <- exp_data %>% replace(is.na(.), 0)

# convert tibble to a matrix for use in heatmap
exp_matrix <- as.matrix(exp_data)

# draw the heatmap
heatmap.2(exp_matrix, scale="column", trace="none", 
    key = FALSE, margin=c(6,8))

```

We can identify two distinct groups of samples within our data with high and low infiltration signatures.  This is a simple example, but hopefully
gives you a flavor of the simplicity and power of working with your data from tag.bio in R.

