---
title: "Getting Started with tagbio.R"
author: "J Ireland, j@tag.bio"
date: "December 10, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started-with-tagbio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## 1 Overview of Using tag.bio with R

The `tagbio.R` package allows users to interact with a tag.bio flux capacitor
(FC) using native R commands.  

### 1.1 Calling tag.bio from R

In the simplest use case, a user may wish to retrieve data from a tag.bio FC
to use in their R application.  In this scenario, the FC is being used as a 
queryable data source, returning results as a dataframe much like they would
get through a database call or simple file upload.  The interaction is one way
with data being pulled from the FC, but no results being returned back to 
tag.bio.  The user is doing almost all of their work from the R environment.
The `tagbio.R` package facilitates this interaction, performing API calls and
formatting results as R data.frame objects.

Even if the user's end goal is to call custom R methods from a tag.bio protocol,
development will typically use this read-only interaction mode until the method
is deemed ready for use.  Once the R method is working in this mode, it is 
straight-forward to then have it be run directly from the tag.bio protocol as 
described below.

### 1.2 Calling R from tag.bio


## 2 Set Up

### 2.1 Connecting to tag.bio from R

`tagbio.R` retrieves data from a tag.bio FC using the built in API.  To connect 
to tag.bio from R, you will need the API URL of the FC you wish to connect to, a 
protocol name (typically "download") and a valid API key for authentication.  

The API URL will be the URL you normally use to access the tag.bio instance 
followed by a "/q".  For example, if you normally access tag.bio at 
<https://fc-skcm.fc.tag.bio>, then the API URL would simply be 
<https://fc-skcm.fc.tag.bio/q>.

To obtain a valid API key you will need to contact <support@tag.bio>.

### 2.2 Configuring tag.bio to Run Custom R Code

...This section will be more involved...

## 3 Example: Calling tag.bio from R

Something about FC

```{r create_fc, echo=TRUE, message=F, warning=F}

library(tagbio)

# set up connection to the FC
fc_hnsc <- FC(name = "fc-hnsc", url = "https://fc-genesig-gateway-azure.dev.tag.bio/fc-svc/fc-hnsc/")

```

```{r create_protocol, echo=TRUE, message=F, warning=F}

# leukocyte infiltration signature genes
leuk_genes <- c("CCL5", "CD19", "CD37", "CD3D", "CD3E", "CD3G", "CD3Z", "CD79A", "CD79B", "CD8A",
    "CD8B1", "IGHG3", "IGJ", "IGLC1", "CD14", "LCK", "LTB", "MS4A1")
leuk_genes <- c("A1CF","A2M")

# define protocol instance
prot_inst <- ProtocolInstance(name = "download", 
    arguments = list(expression = leuk_genes,
                     clinical_categorical_variables = c("clinical.SAMPLE_ID")))

```

```{r get_data, echo=TRUE, message=F, warning=F}

# get

exp_data <- getTagData(fc_hnsc, prot_inst) 
exp_data

```

## 4 Example: Calling R from tag.bio


## 5 Conclusion

## 6 Session Info

